---
title: "Create Tidy Dataset: 500 cities + American Community Survey Data"
date: "12/1/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidycensus)
```

## Description

The purpose of this file is to create a tidy version of the 500 Cities dataset merged onto the ACS data to avoid having to re-run in other programs multiple times since the code takes >5 minutes to run.

The dataset is based on the 2016 data in the 500 cities dataset (The 2015 data doesn't have information on unhealthy behaviors). 

We do not need to link to this RMD file for the website.

```{r, message=FALSE}
# read in dataset, keep records at the census tract level
raw_data <- readRDS("./data/df_500_cities.RDS") %>% 
  janitor::clean_names() %>% 
  filter(geographic_level == "Census Tract", year == 2016)

# clean up data to prepare for a model
tidy_data <- raw_data %>% 
  filter(category == "Unhealthy Behaviors" | short_question_text %in% c("Mental Health", "Health Insurance", "Physical Health")) %>% 
  pivot_wider(id_cols = c(year, state_abbr, state_desc, city_name, geographic_level, geo_location, tract_fips),
              names_from = short_question_text,
              values_from = data_value) %>% 
  janitor::clean_names()

# apply api key
key <- census_api_key("ae24453f5cbe5a92b368876f673eafffbec7b673")

# get list of variables available in the 2015 5-year ACS estimates
# looked through this dataset to identify variables to pull below
vars_acs <- load_variables(2015, "acs5")

# look at concepts
look <- tibble(concepts = unique(sort(pull(vars_acs,concept))))

# for each state in the 500 cities dataset, pull the relevant ACS variables
states <- tibble(states = unique(sort(pull(tidy_data, state_abbr))))

#50 states + DC
# nrow(states)

# for each state, pull the ACS data
# this code works
acs_data <- states %>% 
  # filter(states %in% c("New York", "Alabama")) %>% #for testing code only
  mutate(acs_data = map(states, ~get_acs(geography = "tract", state = .x, 
                                         variable = c("B01001_001", "B01001_002", "B01001_026", #sex
                                                      "B06009_001", "B06009_002", #education
                                                      "B02001_001", "B02001_002"), #race
                                         year = 2015, key = "ae24453f5cbe5a92b368876f673eafffbec7b673"))) %>% 
  unnest(cols = c(acs_data)) %>% 
  pivot_wider(id_cols = c(GEOID, NAME),
              names_from = variable,
              values_from = estimate) %>% 
  mutate(pct_male = round(100*B01001_002/B01001_001, 2),
         pct_female = round(100*B01001_026/B01001_001, 2),
         pct_ltHS = round(100*B06009_002/B06009_001, 2),
         pct_white = round(100*B02001_002/B02001_001, 2)) %>% 
  select(-starts_with("B0"))

# combine ACS data with 500 cities data
tidy_data_acs <- left_join(tidy_data, acs_data, by = c("tract_fips" = "GEOID")) %>% 
  mutate(city_state = paste0(city_name, ", ", state_abbr))

# save data for use in other parts of website
save(tidy_data_acs, file = "./data/tidy_data_acs.Rdata")
```

