---
title: " <b>The Happy Cities:</b> <br> Factors Associated with Geographic Variation in Mental Health Status across 500 Cities in USA"
output: 
  html_document:
    toc: true
    toc_float: true
---


<center>
<figure>
<img src = "images/brain_pic.png" style = "width:60%">

*Image from* [*SARCOIDOSIS UK*](https://www.sarcoidosisuk.org/support/sarcoidosis-and-mental-health/)
</figure>
</center>

## Introduction

Mental illnes is one of the most common medical conditions in the United States. In 2017, 20% of Americans lived with some type of mental illness and 4.6% of Americans had a serious mental illness which substantially interfered with, or limited, one or more major life activities.

For this project we wanted to better understand what behaviors are associated with good and poor mental health and explore the distribution in rates of mental health in the United States. 

## Presentation

```{r echo=FALSE, message = FALSE, warning = FALSE}
library(vembedr)
```

```{r echo=FALSE, message = FALSE, warning=FALSE}
embed_url("https://www.youtube.com/")
```

## Data

This project utilizes data from the [500 cities dataset](https://www.cdc.gov/500cities/index.htm), publicly available for download [here](https://catalog.data.gov/dataset/500-cities-local-data-for-better-health-fc759). The data are based on the [Behavioral Risk Factor Surveillance System (BRFSS)](https://www.cdc.gov/brfss/) survey, which is a national survey of adults' behavior surrounding their health (risk behavior, use of preventative services, chronic conditions). 


## Results

We found that lack of physical activity, smoking and lack health insurance were all associated with poorer mental health. A higher percentage of men, white adults, and adults with less than a high school education were associated with a lower percentage of adults with poorer mental health. 



### Ford Holland | Yongmei Huang | Jessica Lavery | Kylie Riley

#### jfh2150, yh2736, jl4443, kmw2189
