---
title: "Model Building"
author: "Jessica Lavery"
date: "11/17/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidycensus)

# references
# https://walkerke.github.io/tidycensus/articles/basic-usage.html

# set theme for ggplot
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r, message=FALSE}
# read in dataset, keep records at the census tract level
raw_data <- readRDS("./data/df_500_cities.RDS") %>% 
  janitor::clean_names() %>% 
  filter(geographic_level == "Census Tract")

# clean up data to prepare for a model
tidy_data <- raw_data %>% 
  filter(category == "Unhealthy Behaviors" | short_question_text %in% c("Mental Health", "Health Insurance", "Physical Health")) %>% 
  pivot_wider(id_cols = c(year, state_abbr, state_desc, city_name, geographic_level, geo_location, tract_fips),
              names_from = short_question_text,
              values_from = data_value) %>% 
  janitor::clean_names()

# apply api key
key <- census_api_key("ae24453f5cbe5a92b368876f673eafffbec7b673")

# get list of variables available in the 2015 5-year ACS estimates
# looked through this dataset to identify variables to pull below
vars_acs <- load_variables(2015, "acs5")

# look at concepts
look <- tibble(concepts = unique(sort(pull(vars_acs,concept))))

# for each state in the 500 cities dataset, pull the relevant ACS variables
states <- tibble(states = unique(sort(pull(tidy_data, state_abbr))))

#50 states + DC
# nrow(states)

# for each state, pull the ACS data
# this code works
acs_data <- states %>% 
  # filter(states %in% c("New York", "Alabama")) %>% #for testing code only
  mutate(acs_data = map(states, ~get_acs(geography = "tract", state = .x, 
                                         variable = c("B01001_001", "B01001_002", "B01001_026", #sex
                                                      "B06009_001", "B06009_002", #education
                                                      "B02001_001", "B02001_002"), #race
                                         year = 2015))) %>% 
  unnest(cols = c(acs_data)) %>% 
  pivot_wider(id_cols = c(GEOID, NAME),
              names_from = variable,
              values_from = estimate) %>% 
  mutate(pct_male = round(100*B01001_002/B01001_001, 2),
         pct_female = round(100*B01001_026/B01001_001, 2),
         pct_ltHS = round(100*B06009_002/B06009_001, 2),
         pct_white = round(100*B02001_002/B02001_001, 2)) %>% 
  select(-starts_with("B0"))

# combine ACS data with 500 cities data
tidy_data_acs <- left_join(tidy_data, acs_data, by = c("tract_fips" = "GEOID"))
```


## Build model for mental health as an outcome

```{r, message=FALSE}
# look at distribution of data to determine fit of a model
# data are skewed, but not terribly so, will run linear model and then bootstrap for standard errors
ggplot(data = tidy_data_acs, aes(x = mental_health)) +
  geom_histogram(bins = 40) +
  labs(x = "% of Adults with mental health not good for ≥14 days",
       y = "Count", title = "Distribution of the Percent of Adults at the Census-Tract Level with Poor Mental Health",
       caption = "Poor mental health defined as the % of adults who report that their mental health is not good for 14+ days in the past month.")
```

```{r}
# run linear model with unhealthy behaviors as only predictors
mdl_mental_heatlh <- lm(mental_health ~ physical_inactivity + current_smoking + health_insurance + pct_male + pct_ltHS + pct_white, data = tidy_data_acs)

# review model output
# summary(mdl_mental_heatlh)
# mdl_mental_heatlh %>% broom::glance()
# coef(mdl_mental_heatlh)

broom::tidy(mdl_mental_heatlh) %>% 
  mutate(term = case_when(term == "physical_inactivity" ~ "% Without physicial activity",
                          term == "current_smoking" ~ "% Current smokers",
                          term == "health_insurance" ~ "% Without health insurance",
                          term == "pct_male" ~ "%  Male",
                          term == "pct_ltHS" ~ "% With less than high school education",
                          term == "pct_white" ~ "% White",
                          TRUE ~ term),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  knitr::kable(digits = c(2, 2, 2, 2, 2))

```

## Assess model fit

The residuals are evenly dispersed around the predicted values, indicating that there are no obvious problems with the model.

```{r}
# add diagnostics to data frame
diagnostic_mh <- tidy_data_acs %>% 
  modelr::add_residuals(mdl_mental_heatlh) %>% 
  modelr::add_predictions(mdl_mental_heatlh)

# plot diagnostics
ggplot(data = diagnostic_mh, aes(x = pred, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "gray") + 
  labs(x = "Predicted value",
      y = "Residual")
```

## Hypothesis testing

Because the outcome is slightly skewed, instead of relying on traditional frameworks for assessing the significance of each variable, we will bootstrap to estimate the standard errors. 

```{r}
# 1. select bootstrap samples
bootstrap_samples <- tidy_data_acs %>% 
  modelr::bootstrap(n = 10)

# 2. on each bootstrapped sample, run a linear model and extract relevant information 
bootstrap_models <- bootstrap_samples %>% 
  mutate(models = map(strap, ~lm(tmax ~ tmin, data = .x)),
         results = map(models, broom::tidy),
         r2 = map(models, broom::glance)) %>% 
  select(-strap, -models) %>%
  unnest(c(results, r2), names_repair = "universal") %>% 
  select(.id, term, estimate, r.squared) %>% 
  pivot_wider(id_cols = c(.id, r.squared),
              names_from = term,
              values_from = estimate) %>% 
  rename(beta0 = `(Intercept)`, beta1 = "tmin") %>% 
  mutate(logb0b1 = log(beta0 * beta1)) 
```

