---
title: "Report"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(patchwork)
library(broom.mixed)

# set theme for ggplot
theme_set(theme_bw() + theme(legend.position = "bottom"))

# load tidied 500 cities + ACS data
load("./data/tidy_data_acs.Rdata")
```

## **Motivation**

Mental illnes is one of the most common medical conditions in the United States. In 2017 alone, 46.6 million (20%) Americans lived with some type of mental illness and more than 10 million (4.6%) Americans had a serious mental illness which substantially interfered with, or limited, one or more major life activities. Everyone has the potential to develop a mental illness over the course of their life, regardless of age, gender, ethincity, or socioecomonic status. 

Geographic variation of mental health has drawn attention from the health care community. A recent report found that the majority of cities with high rankings for good mental health are located in the Midwest and West, while none of the best cities for mental health are located in the Northeast. Factors that contribute to variation in mental health status are not well understood and need further study. 

Given the high prevalence of mental illness and large disease burden, our group sought to vizualize distributions of mental health across the United States and to explore what behaviors, if any, are signficiantly associated with mental health. 

 
## **Related Work**

We were inspired to utilize the data from the 500 Cities Dataset after seeing how the data was utilized in a prior project for the P8105 class, [The Heart Project](https://alc2279.github.io/p8105_final_project/).


## **Initial Questions**

When we first started brainstorming for this project we were interested in looking at cancer, specifically uterine cancer, as an endpoint. Our inital plan was to look at behaviors associated with uterine cancer and to examine the geographic variation in cancer rates. The 500 Cities dataset includes data on cancer, but after further exploration we discovered that there was cancer variable available which included all cancers except for melanoma, and the group felt that the data was too general to do a meaningful analysis. 

We switched our outcome to mental health because we felt it was a timely and interesting topic, and because the data supported the scope of the project. We continuted with our plan to look at the association of health behaviors with mental health as the outcome and to look at the geographic variation. 


## **Data & Methods**

This project utilizes data from the [500 cities dataset](https://www.cdc.gov/500cities/index.htm), publicly available for download [here](https://catalog.data.gov/dataset/500-cities-local-data-for-better-health-fc759). The data are based on the [Behavioral Risk Factor Surveillance System (BRFSS)](https://www.cdc.gov/brfss/) survey, which is a national survey of adults' behavior surrounding their health (risk behavior, use of preventative services, chronic conditions). Data are summarized using small area estimation methods and are not reported on the individual level, but at the census-tract level and include geographic identifiers such as latitude and longitude, city Federal Information Processing Standards (FIPS) and census tract FIPS. 

The 500 cities data source includes data on health outcomes such as asthma, arthritis, cancer, heart disease, and mental health, along with data on unhealthy behaviors such as drinking, smoking, and obesity. 

The Centers for Disease Control and Prevention’s (CDC) 500 Cities 2018 data was used for this analysis, linked with American Community Survey census tract data. All data is for the population 18 years and over. The 500 Cities Project is to provide city- and census tract-level health data in the largest 500 cities in the US, including chronic disease risk factors, health outcomes, and clinical preventive service use. The primary outcome is self-reported mental health where adults are considered to be in good mental health if they report being in good mental for a majority of days in a given month. Cities were ranked based on the percentage of adults reporting good mental health.

We also merged data from the [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs/data.html) at the census level to the 500 cities dataset. From the ACS data, we obtained demographic information, such as the percent of adults in the census district that were male, white, and had less than a high school education. 

A random effects model was developed to examine the geographic variation across cities. The linear model was adjusted for factors identified a priori; a model-building process was not employed. Factors included the percent of adults without leisure-time physical activity, percent of adults that were current smokers, the percent of adults without health insurance (all from the 500 cities dataset) in addition to the percent male, with less than high school education, and white (from the American Community Survey data). The model is clustered at the city level, accounting for repeated measures at the census tract level within each city. 


## **Exploratory Analysis**

We started by looking at the distribution of mental health across all census districts in the US and averaged the percent of adults with poor mental health for 14 of the past 30 days across all census tracts within a city. We can see that the data are skewed slightly, but that the average percent of adults with poor mental health was `r round(mean(pull(tidy_data_acs %>% drop_na(mental_health), mental_health)), 1)`% (standard deviation `r round(sd(pull(tidy_data_acs %>% drop_na(mental_health), mental_health)), 2)`%).

```{r, message=FALSE}
# look at distribution of data to determine fit of a model
# data are skewed, but not terribly so, will run linear model and then bootstrap for standard errors
ggplot(data = tidy_data_acs %>% drop_na(mental_health), aes(x = mental_health)) +
  geom_histogram(bins = 40) +
  labs(x = "% of Adults with mental health not good for ≥14 days",
       y = "Count", title = "Distribution of the Percent of Adults at the Census-Tract Level with Poor Mental Health",
       caption = "Poor mental health defined as the % of adults who report that their mental health is not good for 14+ days in the past month.")
```

We were also interested in which cities have the highest and lowest mental health, unadjusted for other factors. We can see that overall, there isn't a large difference between cities with the best and worst mental health; all cities had fewer than 25% of adults reporting poor mental health in the past month. Among the top 10 in both directions, cities were fairly close to each other with respect to the percentage of adults with poor mental health. 

```{r}
summaries_by_city <- tidy_data_acs %>% 
  group_by(state_desc, city_name) %>% 
  summarize_if(is.numeric, mean, na.rm = TRUE) %>% 
  mutate(city_state = paste0(city_name, ", ", state_desc)) %>% 
  ungroup()

# cities with best and worst mental health
worst_mh <- summaries_by_city %>% 
  top_n(10, mental_health) %>% 
  ggplot(aes(x = reorder(city_state, -desc(mental_health)), y = mental_health)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the bottom 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

best_mh <- summaries_by_city %>% 
  top_n(-10, mental_health) %>% 
  ggplot(aes(x = reorder(city_state, -desc(mental_health)), y = mental_health)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the top 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

# panel figures
best_mh / worst_mh
```

We looked at the association of health behaviors with mental health and saw that there is a high correlation between many unhealthy behaviors and poorer mental health. 

```{r}

```


## **Results**

### Maps

### Hierarchical linear model for mental health

```{r, include=FALSE, eval=FALSE}
# first pass at model: fixed effects model
# run linear model with unhealthy behaviors as only predictors
mdl_mental_heatlh <- lm(mental_health ~ physical_inactivity + current_smoking + health_insurance + pct_male + pct_ltHS + pct_white, data = tidy_data_acs )

# review model output
# summary(mdl_mental_heatlh)
# mdl_mental_heatlh %>% broom::glance()
# coef(mdl_mental_heatlh)

broom::tidy(mdl_mental_heatlh) %>% 
  mutate(term = case_when(term == "physical_inactivity" ~ "% Without physicial activity",
                          term == "current_smoking" ~ "% Current smokers",
                          term == "health_insurance" ~ "% Without health insurance",
                          term == "pct_male" ~ "%  Male",
                          term == "pct_ltHS" ~ "% With less than high school education",
                          term == "pct_white" ~ "% White",
                          TRUE ~ term),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  knitr::kable(digits = c(2, 2, 2, 2, 2))

# add diagnostics to data frame
diagnostic_mh <- tidy_data_acs %>% 
  modelr::add_residuals(mdl_mental_heatlh) %>% 
  modelr::add_predictions(mdl_mental_heatlh)

# plot diagnostics
# The residuals are evenly dispersed around the predicted values, indicating that there are no obvious problems with the model.
ggplot(data = diagnostic_mh, aes(x = pred, y = resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "gray") + 
  labs(x = "Predicted value",
      y = "Residual")
```

Looking at the distribution of the data at the census-tract level (above), the % of adults with poor mental health for 14+ days out of the last 30 is slightly skewed. Due to the large sample size we will proceed with a linear model and ignore the slight skeweness in the data. 

We can see that a higher percentage of adults without physical activity is associated with better mental health. Smoking and lack health insurance were associated with poorer mental health. A higher percentage of men, white adults, and adults with less than a high school education were associated with a lower percentage of adults with poorer mental health. 

```{r}
# hierarchical model
mdl_mh_random <- nlme::lme(mental_health ~ physical_inactivity + current_smoking + health_insurance + pct_male + pct_ltHS + pct_white, random = ~ 1 | city_state, data = tidy_data_acs %>% drop_na())

# get model output from random effects model
broom.mixed::tidy(mdl_mh_random) %>% 
  filter(effect == "fixed") %>% 
  mutate(term = case_when(term == "physical_inactivity" ~ "% Without physicial activity",
                          term == "current_smoking" ~ "% Current smokers",
                          term == "health_insurance" ~ "% Without health insurance",
                          term == "pct_male" ~ "% Male",
                          term == "pct_ltHS" ~ "% With less than high school education",
                          term == "pct_white" ~ "% White",
                          TRUE ~ term),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  select(-effect, -group, -df, -statistic) %>% 
  knitr::kable(digits = c(2, 3, 3, 2, 2))
```

From the adjusted model, below are the cities with the best and worst mental health.

```{r}
pred <- as_tibble(predict(mdl_mh_random, tidy_data_acs %>% drop_na(), level = 0:1)) %>% 
  group_by(city_state) %>% 
  summarize_if(is.numeric, mean, na.rm = TRUE) %>% 
  ungroup()

# cities with best and worst mental health
worst_mh <- pred %>% 
  top_n(10, predict.city_state) %>% 
  mutate(category = "Worst predicted mental health") %>% 
  arrange(desc(predict.city_state))

best_mh <- pred %>% 
  top_n(-10, predict.city_state) %>% 
  mutate(category = "Best predicted mental health") %>% 
  arrange(predict.city_state)

# comebine into table
bind_rows(worst_mh, best_mh) %>% 
  select(category, city_state, predict.city_state) %>% 
  rename(`Predicted % adults with poor mental health` = predict.city_state) %>% 
  knitr::kable(digits = c(1, 1, 2))
```

## **Discussion**
