---
title: "Report"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(patchwork)
library(broom.mixed)
library(corrplot)

# set theme for ggplot
theme_set(theme_bw() + theme(legend.position = "bottom"))

# load tidied 500 cities + ACS data
load("./data/tidy_data_acs.Rdata")
```

## **Motivation**

Mental illnes is one of the most common medical conditions in the United States. In 2017 alone, 46.6 million (20%) Americans lived with some type of mental illness and more than 10 million (4.6%) Americans had a serious mental illness which substantially interfered with, or limited, one or more major life activities. Everyone has the potential to develop a mental illness over the course of their life, regardless of age, gender, ethincity, or socioecomonic status. 

Geographic variation of mental health has drawn attention from the health care community. A recent report found that the majority of cities with high rankings for good mental health are located in the Midwest and West, while none of the best cities for mental health are located in the Northeast. Factors that contribute to variation in mental health status are not well understood and need further study. 

Given the high prevalence of mental illness and large disease burden, our group sought to vizualize distributions of mental health across the United States and to explore what behaviors, if any, are signficiantly associated with mental health. 

 
## **Related Work**

We were inspired to utilize the data from the 500 Cities Dataset after seeing how the data was utilized in a prior project for the P8105 class, [The Heart Project](https://alc2279.github.io/p8105_final_project/).


## **Initial Questions**

When we first started brainstorming for this project we were interested in looking at cancer, specifically uterine cancer, as an endpoint. Our inital plan was to look at behaviors associated with uterine cancer and to examine the geographic variation in cancer rates. The 500 Cities dataset includes data on cancer, but after further exploration we discovered that there was cancer variable available which included all cancers except for melanoma, and the group felt that the data was too general to do a meaningful analysis. 

We switched our outcome to mental health because we felt it was a timely and interesting topic, and because the data supported the scope of the project. We continuted with our plan to look at the association of health behaviors with mental health as the outcome and to look at the geographic variation. 


## **Data & Methods**

This project utilizes data from the [500 cities dataset](https://www.cdc.gov/500cities/index.htm), publicly available for download [here](https://catalog.data.gov/dataset/500-cities-local-data-for-better-health-fc759). The data are based on the [Behavioral Risk Factor Surveillance System (BRFSS)](https://www.cdc.gov/brfss/) survey, which is a national survey of adults' behavior surrounding their health (risk behavior, use of preventative services, chronic conditions). Data are summarized using small area estimation methods and are not reported on the individual level, but at the census-tract level and include geographic identifiers such as latitude and longitude, city Federal Information Processing Standards (FIPS) and census tract FIPS. 

The 500 cities data source includes data on health outcomes such as asthma, arthritis, cancer, heart disease, and mental health, along with data on unhealthy behaviors such as drinking, smoking, and obesity. 

The Centers for Disease Control and Prevention’s (CDC) 500 Cities 2018 data was used for this analysis, linked with American Community Survey census tract data. All data is for the population 18 years and over. The 500 Cities Project is to provide city- and census tract-level health data in the largest 500 cities in the US, including chronic disease risk factors, health outcomes, and clinical preventive service use. The primary outcome is self-reported mental health where adults are considered to be in good mental health if they report being in good mental for a majority of days in a given month. Cities were ranked based on the percentage of adults reporting good mental health.

We also merged data from the [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs/data.html) at the census level to the 500 cities dataset. From the ACS data, we obtained demographic information, such as the percent of adults in the census district that were male, white, and had less than a high school education. 

A random effects model was developed to examine the geographic variation across cities. We first explored the data and examined the association between availabile health behavior factors and mental health. We then examined the pairwise correlations between the health behavior factors and found (perhaps not surprisingly) a high correlation among most factors. Given this, we were unable to concurrently examine the association of many health behaviors at once. We were primarily interested in the assocation between physical inactivity and mental health, so we included physical inactivity. The linear model was further adjusted for factors identified a priori from the American Community Survey data and included the percent male, with less than high school education, and white. The model is clustered at the city level, accounting for repeated measures at the census tract level within each city. 


## **Exploratory Analysis**

We started by looking at the distribution of mental health across all census districts in the US and averaged the percent of adults with poor mental health for 14 of the past 30 days across all census tracts within a city. We can see that the data are skewed slightly, but that the average percent of adults with poor mental health was `r round(mean(pull(tidy_data_acs %>% drop_na(mental_health), mental_health)), 1)`% (standard deviation `r round(sd(pull(tidy_data_acs %>% drop_na(mental_health), mental_health)), 2)`%).

```{r, message=FALSE, fig.height=3}
# look at distribution of data to determine fit of a model
# data are skewed, but not terribly so, will run linear model and then bootstrap for standard errors
ggplot(data = tidy_data_acs %>% drop_na(mental_health), aes(x = mental_health)) +
  geom_histogram(bins = 40) +
  labs(x = "% of Adults with mental health not good for ≥14 days",
       y = "Count", title = "Distribution of the Percent of Adults at the Census-Tract Level with Poor Mental Health",
       caption = "Poor mental health defined as the % of adults who report that their mental health is not good for 14+ days in the past month.")
```

We were also interested in which cities have the highest and lowest mental health, unadjusted for other factors. We can see that overall, there isn't a large difference between cities with the best and worst mental health; all cities had fewer than 25% of adults reporting poor mental health in the past month. Among the top 10 in both directions, cities were fairly close to each other with respect to the percentage of adults with poor mental health. 

```{r}
summaries_by_city <- tidy_data_acs %>% 
  group_by(state_desc, city_name) %>% 
  summarize_if(is.numeric, mean, na.rm = TRUE) %>% 
  mutate(city_state = paste0(city_name, ", ", state_desc)) %>% 
  ungroup()

# cities with best and worst mental health
worst_mh <- summaries_by_city %>% 
  top_n(10, mental_health) %>% 
  ggplot(aes(x = reorder(city_state, -desc(mental_health)), y = mental_health)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the bottom 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

best_mh <- summaries_by_city %>% 
  top_n(-10, mental_health) %>% 
  ggplot(aes(x = reorder(city_state, -desc(mental_health)), y = mental_health)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the top 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

# panel figures
best_mh / worst_mh
```

#### **Maps**

We created several maps to visualize the distribution of mental health outcomes across the United States. The nationwide scatter plot map is rendered using `plotly's` integration with mapbox and shows the reported percentage of poor mental health by census tract. This map was originally intended as a census-level choropleth, but because of loading-performance issues, we decided to display it as a scatter plot and supplement with smaller-scale choropleths.

The choropleths for New York, San Francisco, and Chicago use geographical geometries from the `tidycensus` package that were joined to the 500 Cities data by census tract FIPS code. These maps display the census level tracts colored by the average percent of poor mental health for each district.

The scatter plot shows a relatively even distribution of mental scores across the country. The most apparent trends are slightly higher percentages of poor mental health in the northeast and lower values along the west coast. This observation is reinforced by our exploratory analysis, which found that several of the cities with the poorest mental health outcomes are distributed on the east coast, while several of the happiest cities in the country are in California.

The choropleths provide a more detailed picture of mental health in the cities presented. Of the three maps, Chicago seems to have the most heterogeneity in mental health, with pockets of poor outcomes in the central and southern districts. San Francisco appears to be the happiest of the three, and New York has noticeable differences between northern and southern census tracts. 

The happiest district in Manhattan is a few blocks in the Upper East Side along Central Park, where only 6.5% of respondents report poor mental health.

**Associations between mental health and other health behviors at the census-tract level**

```{r, message=FALSE}
# physical health
phys_health <- ggplot(data = summaries_by_city, aes(x = physical_inactivity, y = mental_health)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% No leisure time physical activity",
       title = "Physical health")

# binge drinking
binge_drinking <- ggplot(data = summaries_by_city, aes(x = binge_drinking, y = mental_health)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% Binge drinking",
       title = "Binge drinking")

# health insurance
health_ins <- ggplot(data = summaries_by_city, aes(x = health_insurance, y = mental_health)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% No health insurance",
       title = "Health insurance")

# current smoking
current_smk <- ggplot(data = summaries_by_city, aes(x = current_smoking, y = mental_health)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% Current Smokers",
       title = "Smoking status")

# obesity
obesity <- ggplot(data = summaries_by_city, aes(x = obesity, y = mental_health)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% Obese",
       title = "Obesity")

# sleep_7_hours
sleep <- ggplot(data = summaries_by_city, aes(x = obesity, y = sleep_7_hours)) +
  geom_point() +
  geom_smooth() +
  labs(y = "% Poor Mental Health",
       x = "% Insufficient Sleep",
       title = "Insufficient Sleep")

(phys_health + binge_drinking) / (obesity + sleep) / (health_ins + current_smk)
``` 

Definitions: 

* Poor mental health defined as the percent of adults who report having poor mental health for 14 of the last 30 days.

* Physical health defined as the percent of adults who report not participating in any physical activities or exercises such as running, calisthenics, golf, gardening, or walking for exercise. 

* Binge drinking defined as the percent of adults who report having five or more drinks (men) or four or more drinks (women) on an occasion in the past 30 days.

* Obesity defined as the percent of adults whose BMI is 30 kg/m2 or greater. 

* Insufficient sleep defined as the percent of adults who report usually getting insufficient (<7 hours) sleep. 

* Lack of health insurance was defined as the percentage of adults (aged 18–64 years; i.e. those not eligible for Medicare) who report having current health insurance or having no current health insurance

* Current smoking defined as the percent of adults who report smoking ≥100 cigarettes in their lifetime and currently smoke every day.

Ideally, we would adjust for all of these unhealthy behaviors in a single model. However, we suspected that many of them were correlated and would introduce collinearity to the model. Before proceeding to the model, we looked at the correlations between the pairwise combinations of the above factors. We can see that the behaviors are very strongly positively correlated with each other, with the exception of binge drinking which is negatively correlated with the other factors.

```{r}
cor_mtx <- cor(tidy_data_acs %>% 
                 select_if(is.numeric) %>% 
                 select(-year, -mental_health, -physical_health, -starts_with("pct")), 
               use = "pairwise.complete.obs", 
               method = "spearman")
    
corrplot(cor_mtx,  type = "upper",order = "hclust")
```

## **Hierarchical Model Results**

Looking at the distribution of the data at the census-tract level (above), the % of adults with poor mental health for 14+ days out of the last 30 is slightly skewed. Due to the large sample size we will proceed with a linear model and ignore the slight skeweness in the data. 

From the model, we can see that a higher percentage of adults without physical activity is associated with a higher percentage of adults with poor mental health, after adjusting for sex, race and education. A higher percentage of men and adults with less than a high school education were associated with a higher percentage of adults with poorer mental health, where as a higher percentage of white adults in the census district was associated with slightly better mental health; this effect was statistically significant, though likely not practically significant. 

```{r}
# hierarchical model
mdl_mh_random <- nlme::lme(mental_health ~ physical_inactivity + pct_male + pct_ltHS + pct_white, random = ~ 1 | city_state, data = tidy_data_acs %>% drop_na())

# get model output from random effects model
broom.mixed::tidy(mdl_mh_random) %>% 
  filter(effect == "fixed") %>% 
  mutate(term = case_when(term == "physical_inactivity" ~ "% Without physicial activity",
                          term == "current_smoking" ~ "% Current smokers",
                          term == "health_insurance" ~ "% Without health insurance",
                          term == "pct_male" ~ "% Male",
                          term == "pct_ltHS" ~ "% With less than high school education",
                          term == "pct_white" ~ "% White",
                          TRUE ~ term),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  select(-effect, -group, -df, -statistic) %>% 
  knitr::kable(digits = c(2, 3, 3, 2, 2))
```

We formally compared the random effects model to the fixed effects model, and found that the random effect of census tract was statistically signficiant, indicating that there is significant variation in patterns of mental health across cities in the United States. 

```{r}
# compare to fixed effects model
mdl_mh_fixed <- lm(mental_health ~ physical_inactivity + pct_male + pct_ltHS + pct_white, data = tidy_data_acs %>% drop_na())

comparison <- anova(mdl_mh_random,mdl_mh_fixed)
```

From the adjusted hierarchical model, we can look at the distribution of the predicted values for mental health (normal by design of the model) and note that the percent of adults with poor mental health for 14+ days of the past month ranges from xx to xx. 

```{r}
# get predicted values from the hierarchical model
pred <- as_tibble(predict(mdl_mh_random, tidy_data_acs %>% drop_na(), level = 0:1)) %>% 
  group_by(city_state) %>% 
  summarize_if(is.numeric, mean, na.rm = TRUE) %>% 
  ungroup() 

# histogram of predicted values
ggplot(data = pred, aes(x = predict.city_state)) + 
  geom_histogram(bins = 40) +
  labs(x = "Predicted % of Adults with mental health not good for ≥14 days",
       y = "Count", title = "Distribution of the Predicted Percent of Adults at the Census-Tract Level with Poor Mental Health",
       caption = "Poor mental health defined as the % of adults who report that their mental health is not good for 14+ days in the past month.")
```

Below are the cities with the best and worst mental health.

```{r}
# cities with best and worst mental health
worst_mh <- pred %>% 
  top_n(10, predict.city_state) %>% 
  ggplot(aes(x = reorder(city_state, -desc(predict.city_state)), y = predict.city_state)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the bottom 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

best_mh <- pred %>% 
  top_n(-10, predict.city_state) %>% 
  ggplot(aes(x = reorder(city_state, -desc(predict.city_state)), y = predict.city_state)) +
  geom_bar(stat = "identity") +
  labs(title = "Cities with the top 10 mental health in the US",
       x = "",
       y = "% of Adults with mental health not good for ≥14 days") +
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip()

# panel figures
best_mh / worst_mh
```

## **Discussion**

Mental health includes emotional, psychological, and social well-being. Our analysis shows that one out of eight people living in big cities was not feeling well in mental health. People from the west, such as California, had better self-reported mental health than people living in the east. No leisure-time physical activity, insufficient sleep, being obese and being smoking were strongly associated with poor mental health. Variations of poor mental health across 500 cities were found in the random-effect model. Male, lower level of education, and being a minority is associated with a higher likelihood of poor mental health. After adjusting for measured confounders, people from West coast cities were still happier than people living on the East coast. 

Good mental health has been considered as an important factor that allows one to flourish and fully enjoy life. The geographic variation of mental health across the USA is of great interest in many perspectives. First, it provides insights for health policymakers into the allocation of mental health care services. It is obvious that health providers in the East coast area need to provide more mental health counseling for patients. Annual check-ups can be a great opportunity to reach out to a large number of city residers. Secondly, at the individual level, we identified some health behavior factors associated with poor mental health, including limited leisure-time physical activity, the deprivation of sleep, obesity, and smoking. Undoubtedly, physical health and mental health are mutually influenced by one another. They are inseparable. The intervention at the individual level should take a whole picture of one's physical and mental health. Some social factors (such as gender, education, and race)  also have impacts on poor mental health that are hard to intervene on for the manipulability of social factors is limited. Health providers need to pay special attention to patients who are under a high level of stress, such as males, lower levels of education, and minorities.   
 
The strength of our analysis includes the visualization of geographic variation of poor mental health across 500 biggest cities in the USA, covering a substantial of Americans. And, the multilevel analysis allows the testing of heterogeneity of mental health across cities and generates predicted proportion of people with poor mental health at each city after accounting for measured confounders. But, we need to be cautious about the interpretation of ecologic study at the individual level. Our analysis unit is census tract, clustering at the city level. Association found at the neighborhood level might not apply to each individual in the study. For example, a minority male with lower education working as a bus driver may have a happier life than a white male working at the wall street for the latter is, in general, has a much intensive lifestyle with a high risk of being laid-off. Last but not least, our group was discussing the strategy of constructing a composite index to handle the collinearity of health-related factors in the multivariable regression model. For the interest of time, we did not pursue such an analysis. For the future study, it may be worthwhile to build on a health behavior index based on the weights calculated from the logistic regression model via the cross-validation process. 

In conclusion, this "happy cities" project using 500 Cities Data is a meaningful analysis that provides insight into the mental health at the neighborhood level. Good mental health affects our daily life on how we think, feels, and act.  This is an important but maybe an understudied field. 
